---
title: 'Biostatistics II: K-means, Hierarchical clustering'
date: "`r doc_date()`"
output: 
    BiocStyle::html_document:
        toc: true
        highlight: tango
---

```{r lib setup, include=FALSE}
library(ggplot2)
library(plotly)
library("ggfortify")
```

## Materials 

* Course materials are [here.](http://imsieweb.uni-koeln.de/bioinformatik/lehre/biostatistik/)
* Login: CB   
* Password: IMSB

The list of exercises if [here.](./)

# Exercises: K-means, Hierarchical clustering

## Exercise 1

```{r}
require(graphics)

load("../exercises_pca_kmeans/halfmoon.RData")
halfmoon = as.data.frame(halfmoon)

plot(halfmoon[, c("x", "y")])
plot(halfmoon[ , c("x", "y")], col = as.factor(halfmoon[,c("labels")]))
```

```{r, eval = FALSE}
# be careful not tu use the whole data.frame halfmoon for plotting or clustering,
#   as it has 3 columns: x, y, labels
# see what happens if you plot it:
plot(halfmoon)
```

## Exercise 2

```{r, eval = TRUE}
km2 <- kmeans(halfmoon[, c("x", "y")], 2)
plot(halfmoon[, c("x", "y")], col = km2$cluster)
```


## Exercise 3-4

```{r, eval = TRUE}
head(halfmoon)

dist_hmoon <- dist(halfmoon[, c("x","y")], 
                   method = "euclidean")

hc_complete <- hclust(dist_hmoon, method = "complete")
hc_ave      <- hclust(dist_hmoon, method = "average")
hc_single   <- hclust(dist_hmoon, method = "single")
```

Plot the trees for each method:
```{r}
plot(hc_complete)
plot(hc_ave)
plot(hc_single)

```

Plot the roots of the trees (only 2 clusters):
```{r}
root_hc_complete <- cutree(hc_complete, k = 2)
root_hc_ave <- cutree(hc_ave, k = 2)
root_hc_single <- cutree(hc_single, k = 2)

plot(root_hc_single)
plot(root_hc_complete)
plot(root_hc_ave)
```

Plot the 

```{r}
#check that labels are correctly mapped to the data points
stopifnot(names(halfmoon[, c("labels")]) == names(root_hc_complete))
table(halfmoon[, c("labels")] == root_hc_complete - 1)

plot(halfmoon[, c("x","y")], col = as.factor(root_hc_complete))
plot(halfmoon[, c("x","y")], col = as.factor(root_hc_ave))
plot(halfmoon[, c("x","y")], col = as.factor(root_hc_single))

```

```{r, eval = FALSE}
# see what happens if you accidentally feed true labels into the clustering function
dist_hmoon_wrong <- dist(halfmoon, 
                   method = "euclidean")
hc_single2   <- hclust(dist_hmoon_wrong, method = "single")
```

## Exercise 6

Here we want to implement k-median or k-medoid clustering, and try it on the halfmoon data.

We first write the main function, that calculated centers and labels
```{r}
kmedian <- function(data, k)
{   
    # k-median clustering for two-dimentional vectors
    xx <- as.matrix(data)
    m <- as.integer(nrow(xx))
    stopifnot(m>0)
    stopifnot(is.numeric(xx))
    stopifnot(c("x", "y") %in% colnames(xx))
    
    # initialize k centers
    centersIDs_0 <- sample(c(1:nrow(xx)), size = k, replace = FALSE)
    centers_old <- xx[centersIDs_0, c("x", "y")]
    centers <- centers_old
    
    update_happened = 1
    n <- 0 # number of iterations the algorithm took
    while (update_happened == 1)
    {
        # calculate labels
        labels <- apply(xx[, c("x", "y")], 
                        1, 
                        FUN = function(z) get_label(z, centers = centers))
        # redifine centers
        centers_new <- update_centers(xx, labels)
        
        if (identical(centers, centers_new)){   
            update_happened <- 0   
        } else {   
            centers <- centers_new   
            n <- n + 1
        }
    }
    
    #plot(xx, col = as.factor(labels))
    
    res <- list(labels, centers, n)
    return(res)
    
}
```

And to make the code easier to read, we use a couple of gelper functions:
```{r}
manhattan.dist <- function(x1, x2) 
{   
    # manhattan dist between 2 vectors
    return(sum(abs(x1 - x2)))
}

get_label <- function(current_point, centers = centers)
{
    # find distance to the centers from a current point (manhattan)
    dd <- apply(centers, 1, FUN = function(z)manhattan.dist(z, current_point))
    # label from the center with minimal distance
    current_label <- names(dd)[which(dd == min(dd))]
    return(current_label)
}

update_centers <- function(xx, labels)
{
    # calculates medians by component 
    # (not a point from the sets)
    stopifnot(names(labels) == rownames(xx))
    new_center_list <- by(xx, 
                         INDICES = as.factor(labels), 
                         FUN = function(z)
                         {c(median(z$x), median(z$y))})
    new_centers <- do.call("rbind",new_center_list)
    rownames(new_centers) <- paste0("cent_",c(1:nrow(new_centers)))
    colnames(new_centers) <- c("x", "y")
    return(new_centers)
}
```

Now, we finally apply our k-median clustering to the halfmoon data:
```{r}
####################################
load("../exercises_pca_kmeans/halfmoon.RData")
labels_and_centers <- kmedian(data = halfmoon[, c("x", "y")], k = 3)

plot(halfmoon[, c("x", "y")], col = as.factor(labels_and_centers[[1]]))


```
